---
title: "First Attempts at Modeling Unconditional Mean Structure"
author: "Logan Harris"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
library(FITfileR)
library(magrittr)
library(dplyr)
library(tseries)
library(fGarch)
library(lubridate)
library(readr)
library(data.table)
library(janitor)
library(bit64)
library(stringr)
library(splines)
library(ggplot2)
library(rugarch)
```

# Simple removal of trend using ARMA 

```{r}
recs <- records(readFitFile("example_files/Lunch_Gravel_Ride.fit")) 
## recs <- records(readFitFile("example_files/Afternoon_Gravel_Ride.fit")) 
hr_data <- do.call(rbind, lapply(recs, function(x) x %>% select(timestamp, heart_rate))) %>%
  arrange(timestamp) %>%
  mutate(time_diff = c(NA, diff(timestamp))) %>%
  dplyr::filter(!is.na(time_diff))  %>%
  mutate(brk = cumsum(time_diff > 15) + 1) 

## %>% mutate(heart_rate = 60000 / heart_rate)


s_fit <- arma(hr_data$heart_rate, order = c(3, 3))

vals <- as.numeric(fitted(s_fit))

hr_data$vals <- vals
hr_data$centered_hr <- hr_data$heart_rate - vals

hr_data %<>%
  dplyr::filter(!is.na(centered_hr))

ggplot(hr_data, aes(x = timestamp, y = vals, group = brk)) +
  geom_line()

ggplot(hr_data, aes(x = timestamp, y = heart_rate, group = brk)) +
  geom_line()


ggsave(file = "save_test.pdf", width = 12, height = 2, dpi = 300)

ggplot(hr_data, aes(x = timestamp, y = centered_hr, group = brk)) +
  geom_line()

## try and get super smooth trend

span <- 1 / (as.numeric(max(hr_data$timestamp) - min(hr_data$timestamp)) / .25)

fit <- loess(heart_rate ~ as.numeric(timestamp), data = hr_data, span = span)

plot.ts(fit$fitted)
plot.ts(hr_data$heart_rate - fit$fitted)


ugarchfit(spec = ugarchspec(mean.model = list(armaOrder = c(1, 1)),
                                         variance.model = list(garchOrder = c(1, 1))), data = (hr_data$heart_rate - fit$fitted))
```

- Main take away here is that this is not sufficient and that ARMA predictions are not smoothed like we would like from a kalman filter

# BIC selects overly complex model for natural spline

```{r}
bics <- numeric(100)
for (i in 1:100) {
  
  bics[i] <- BIC(lm(heart_rate ~ bs(timestamp, df = 100 + i), data = hr_data))
  
}

plot(1:100, bics)

## Methods of cross validation and types of splines
## ns, bs, penalized
## loocv, gcv, 10-fold (expect to fail), bootstrapped validation, 10 fold intermixed cross validation

group_assignment <- (1:nrow(hr_data) %% 5) + 1
table(group_assignment)
mses <- length(unique(group_assignment))

for (i in 1:length(unique(group_assignment))) {
  
  mod <- lm(heart_rate ~ bs(timestamp, df = 230), data = hr_data[group_assignment != i,])
  preds <- predict(mod, hr_data[group_assignment == i,])
  print(i)
  mses[i] <- mean((preds - unlist(hr_data[group_assignment == i,"heart_rate"]))^2)
  
  
}

sum((mses * table(group_assignment))) / nrow(hr_data)
```


# Simple removal of trend using smoothing

```{r}
res <- gsm(hr_data$heart_rate ~ as.numeric(hr_data$timestamp), method = "GCV")
plot(as.numeric(hr_data$timestamp), fitted(res), type = "l")
plot.ts(hr_data$heart_rate)
vals <- as.numeric(fitted(res))

hr_data$vals <- vals
hr_data$centered_hr <- hr_data$heart_rate - vals

ggplot(hr_data, aes(x = timestamp, y = vals, group = brk)) +
  geom_line()

ggplot(hr_data, aes(x = timestamp, y = heart_rate, group = brk)) +
  geom_line()

ggplot(hr_data, aes(x = timestamp, y = centered_hr, group = brk)) +
  geom_line()

## model sigma t
sfit <- ugarchfit(spec = ugarchspec(mean.model = list(armaOrder = c(0, 0)),
                                         variance.model = list(garchOrder = c(1, 1))), data = hr_data$centered_hr)

plot.ts(sfit@fit$sigma)

ys <- sfit@fit$sigma
xs <- 60000 / hr_data$heart_rate

dat_pred <- bind_cols(ys, xs, hr_data$timestamp)
colnames(dat_pred) <- c("Y", "X", "time")

plot(dat_pred$X, log(dat_pred$Y))
mod <- lm(log(Y) ~ X, dat_pred)

library(visreg)
visreg::visreg(mod, "X")
## visreg::visreg(mod, "time")

summary(mod)
exp(predict(mod, newdata = data.frame(X = 140)))
exp(predict(mod, newdata = data.frame(X = 51)))

plot(xs, log(ys))
lines(mod$model$X, mod$fitted.values, col = "red")
```





## Activity Details

- Preps the information for the activities to be read in

```{r}
activity_details <- fread("./strava_new/activities.csv") %>%
  clean_names()

activity_details %<>%
  dplyr::mutate(activity_date = mdy_hms(activity_date), distance = as.numeric(distance)) %>%
  dplyr::filter(activity_type %in% c("Ride", "Virtual Ride")) %>%
  dplyr::filter(activity_date <= ymd("2022-02-01") & activity_date >= ymd("2021-08-15")) %>%
  ## dplyr::mutate(year = year(activity_date), month = month(activity_date), day = day(activity_date)) %>%
  ## dplyr::group_by(year, month, day) %>%
  dplyr::arrange(activity_date) %>%
  dplyr::mutate(time_diff = c(NA, diff(as.numeric(activity_date))),
                last_time = dplyr::lag(elapsed_time),
                time_diff = time_diff - last_time,
                new_group = cumsum(ifelse(time_diff > 3600 & !is.na(time_diff), 1, 0))) %>%
  ## dplyr::ungroup() %>%
  dplyr::group_by(new_group) %>%
  dplyr::mutate(occurance = 1:n(), total_distance = sum(distance), moc = max(occurance)) %>%
  dplyr::ungroup()  %>%
  dplyr::filter(distance > 16 | moc >= 2)
```


## Use spline to remove mean trend

- mehhh

```{r}
preds <- numeric(nrow(activity_details))
omega <- numeric(nrow(activity_details))
alpha1 <- numeric(nrow(activity_details))
beta1 <- numeric(nrow(activity_details))
dates <- numeric(nrow(activity_details))
slopes <- numeric(nrow(activity_details))
## 163, 177
aerobic <- numeric(nrow(activity_details))
thresh <- numeric(nrow(activity_details))
act_type <- activity_details$activity_type

i <- 1
while (i <= nrow(activity_details)) {
  
  recs <- records(readFitFile(paste0("./strava/", str_remove(activity_details$filename[i], ".gz"))))
      
  if (!("list" %in% class(recs))) {recs <- list(recs)}
  
  hr_data <- do.call(rbind, lapply(recs, function(x) x %>% dplyr::select(timestamp, heart_rate))) %>%
    dplyr::arrange(timestamp)
  
  print(i)
  
  while (i + 1 < nrow(activity_details) && activity_details$occurance[i + 1] != 1) {
    
    i <- i + 1
    
    recs <- records(readFitFile(paste0("./strava/", str_remove(activity_details$filename[i], ".gz"))))
      
    if (!("list" %in% class(recs))) {recs <- list(recs)}
  
    hr_data <- hr_data %>%
      bind_rows(do.call(rbind, lapply(recs, function(x) x %>% dplyr::select(timestamp, heart_rate)))) %>%
      dplyr::arrange(timestamp)
    
  }
  
  hr_data <- hr_data %>%
    mutate(time_diff = c(NA, diff(timestamp))) %>%
    dplyr::filter(!is.na(time_diff))  %>%
    mutate(brk = cumsum(time_diff > 15) + 1) %>%
    dplyr::filter(heart_rate > 40)

  
  tryCatch({
    
    vals <- numeric()
    for (j in 1:length(unique(hr_data$brk))) {
      
     
      
      tmp <- hr_data %>%
        dplyr::filter(brk == j)
      
      s_fit <- lm(heart_rate ~ ns(timestamp, df = round(sum(as.numeric(diff(tmp$timestamp))) / 60)), data = tmp)
      vals <- c(vals, s_fit$fitted.values)
    
    }

      hr_data$vals <- vals
      hr_data$centered_hr <- hr_data$heart_rate - vals
      
      hr_data %<>%
        dplyr::filter(!is.na(centered_hr))
      
      ride_garch <- garchFit(centered_hr~garch(1,0), data = hr_data, trace = FALSE)
      mod <- lm(log(ride_garch@sigma.t) ~ heart_rate, data = hr_data)
      slopes[i] <- coef(mod)[2]
      aerobic[i] <- predict(mod, data.frame(heart_rate = 140))
      dates[i] <- as.numeric(date(min(hr_data$timestamp)))
      omega[i] <- ride_garch@fit$coef[2]
      alpha1[i] <- ride_garch@fit$coef[3]
      ## beta1[i] <- ride_garch@fit$coef[4]
      
      lm()
      
    },
           error = function(e) {
             preds[i] <- NA
             dates[i] <- NA
             }
  )
  
  i <- i + 1
  
}

dates <- dates[!(omega == 0)]
act_type <- act_type[!(omega == 0)]
slopes <- slopes[!(omega == 0)]
aerobic <- aerobic[!(omega == 0)]
thresh <- thresh[!(omega == 0)]
alpha1 <- alpha1[!(alpha1 == 0)]
omega <- omega[!(omega == 0)]
## beta1 <- beta1[!(beta1 == 0)]

plot(dates, alpha1, type = "l")
abline(lm(alpha1[act_type != "Ride"] ~ dates[act_type != "Ride"]), col = "red")

plot(dates, alpha1, type = "p", col = c("blue", "red")[1 + (act_type == "Ride")])

plot(dates, aerobic, type = "l")
```

- Still way too noisy


## Example of modeling sigma_t with mean (ARMA-GARCH)

```{r}
## Exploring ideas
recs <- records(readFitFile("example_files/Afternoon_Gravel_Ride.fit")) 
## recs <- records(readFitFile("Lunch_Gravel_Ride.fit")) 
hr_data <- do.call(rbind, lapply(recs, function(x) x %>% select(timestamp, heart_rate))) %>%
  arrange(timestamp) %>%
  mutate(time_diff = c(NA, diff(timestamp))) %>%
  dplyr::filter(!is.na(time_diff))  %>%
  mutate(brk = cumsum(time_diff > 15) + 1) %>%
  mutate(heart_rate = 60000 / heart_rate)

hr_data %<>%
  group_by(brk) %>%
  mutate(idx = 1:n()) %>%
  ungroup() %>%
  dplyr::filter(idx > 10)

sfit <- ugarchfit(spec = ugarchspec(mean.model = list(armaOrder = c(1, 0)),
                                         variance.model = list(garchOrder = c(1, 10))), data = hr_data$heart_rate)

plot.ts(sfit@fit$sigma)
ys <- sfit@fit$sigma
xs <- 60000 / hr_data$heart_rate

dat_pred <- bind_cols(ys, xs)
colnames(dat_pred) <- c("Y", "X")
mod <- lm(Y ~ X + I(X^2), dat_pred)
predict(mod, newdata = data.frame(X = 150))

plot(xs, ys)
lines(xs, mod$fitted.values, col = "red")
```


## Model with ARMA-GARCH for longitudinal records

```{r}
i <- 1
while (i <= nrow(activity_details)) {
  
  recs <- records(readFitFile(paste0("./strava/", str_remove(activity_details$filename[i], ".gz"))))
      
  if (!("list" %in% class(recs))) {recs <- list(recs)}
  
  hr_data <- do.call(rbind, lapply(recs, function(x) x %>% dplyr::select(timestamp, heart_rate))) %>%
    dplyr::arrange(timestamp)
  
  ## hr_data <- do.call(rbind, lapply(recs, function(x) x %>% dplyr::select(timestamp, heart_rate))) 
  
  print(i)
  
  while (i + 1 < nrow(activity_details) && activity_details$occurance[i + 1] != 1) {
    
    i <- i + 1
    
    recs <- records(readFitFile(paste0("./strava/", str_remove(activity_details$filename[i], ".gz"))))
      
    if (!("list" %in% class(recs))) {recs <- list(recs)}
  
    hr_data <- hr_data %>%
      bind_rows(do.call(rbind, lapply(recs, function(x) x %>% dplyr::select(timestamp, heart_rate)))) %>%
      dplyr::arrange(timestamp)
    
  }
  
  min_time <- min(hr_data$timestamp)
  max_time <- max(hr_data$timestamp)
  all_times <-  seq(min_time, max_time, by = 1)
  all_times <- all_times[!(all_times %in% hr_data$timestamp)]
  
  if (length(all_times) > 0) {
    
   all_ts <- data.frame(timestamp = all_times, heart_rate = NA)  
   hr_data %<>% full_join(all_ts) %>%
    arrange(timestamp) %>%
    mutate(bef = heart_rate,
           aft = heart_rate) %>%
    tidyr::fill(bef, .direction = "down") %>%
    tidyr::fill(aft, .direction = "up") %>%
    mutate(group = cumsum(!is.na(heart_rate))) %>%
    group_by(group) %>%
    mutate(bef_weight = n():1) %>%
    ungroup() %>%
    arrange(desc(timestamp)) %>%
    mutate(group = cumsum(!is.na(heart_rate))) %>%
    group_by(group) %>%
    mutate(aft_weight = n():1) %>%
    ungroup() %>%
    arrange(timestamp) %>%
    mutate(heart_rate = ifelse(is.na(heart_rate), (bef*bef_weight + aft*aft_weight) / (bef_weight + aft_weight), heart_rate))
    
  }

  
  hr_data <- hr_data %>%
    mutate(heart_rate = 60000 / heart_rate) %>%
    mutate(tp = 1:n()) %>%
    dplyr::filter(tp > 600)

  
  tryCatch({


    sfit <- ugarchfit(spec = ugarchspec(mean.model = list(armaOrder = c(1, 1)),
                                         variance.model = list(garchOrder = c(1, 1))), data = hr_data$heart_rate)

    plot.ts(sfit@fit$sigma)
    ys <- sfit@fit$sigma
    xs <- 60000 / hr_data$heart_rate
    
    dat_pred <- bind_cols(ys, xs)
    colnames(dat_pred) <- c("Y", "X")
    mod <- lm(Y ~ X + I(X^2), dat_pred)
    aerobic[i] <- predict(mod, newdata = data.frame(X = 150))
    dates[i] <- as.numeric(date(min(hr_data$timestamp)))
      
    },
           error = function(e) {
             dates[i] <- NA
            }
  )
  
  i <- i + 1
  
}

saved <- list(dates, act_type, aerobic)

dates <- dates[!(aerobic == 0)]
act_type <- act_type[!(aerobic == 0)]
aerobic <- aerobic[!(aerobic == 0)]

plot(dates, aerobic, type = "l")
abline(lm(aerobic[act_type != "Ride"] ~ dates[act_type != "Ride"]), col = "red")

plot(dates, aerobic, type = "p", col = c("blue", "red")[1 + (act_type == "Ride")])
```

## Splines with GCV

```{r}
dates <- numeric(nrow(activity_details))
aerobic <- numeric(nrow(activity_details))
act_type <- activity_details$activity_type

i <- 1
while (i <= nrow(activity_details)) {
  
  recs <- records(readFitFile(paste0("./strava_new/", str_remove(activity_details$filename[i], ".gz"))))
      
  if (!("list" %in% class(recs))) {recs <- list(recs)}
  
  hr_data <- do.call(rbind, lapply(recs, function(x) x %>% dplyr::select(timestamp, heart_rate))) %>%
    dplyr::arrange(timestamp) %>%
      dplyr::filter(heart_rate > 30)
  
  ## hr_data <- do.call(rbind, lapply(recs, function(x) x %>% dplyr::select(timestamp, heart_rate))) 
  
  print(i)
  
  while (i + 1 < nrow(activity_details) && activity_details$occurance[i + 1] != 1) {
    
    i <- i + 1
    
    recs <- records(readFitFile(paste0("./strava_new/", str_remove(activity_details$filename[i], ".gz"))))
      
    if (!("list" %in% class(recs))) {recs <- list(recs)}
  
    hr_data <- hr_data %>%
      bind_rows(do.call(rbind, lapply(recs, function(x) x %>% dplyr::select(timestamp, heart_rate)))) %>%
      dplyr::arrange(timestamp) %>%
      dplyr::filter(heart_rate > 30)
    
  }
  
  min_time <- min(hr_data$timestamp)
  max_time <- max(hr_data$timestamp)
  all_times <-  seq(min_time, max_time, by = 1)
  all_times <- all_times[!(all_times %in% hr_data$timestamp)]
  
  if (length(all_times) > 0) {
    
   all_ts <- data.frame(timestamp = all_times, heart_rate = NA)  
   hr_data %<>% full_join(all_ts) %>%
    arrange(timestamp) %>%
    mutate(bef = heart_rate,
           aft = heart_rate) %>%
    tidyr::fill(bef, .direction = "down") %>%
    tidyr::fill(aft, .direction = "up") %>%
    mutate(group = cumsum(!is.na(heart_rate))) %>%
    group_by(group) %>%
    mutate(bef_weight = n():1) %>%
    ungroup() %>%
    arrange(desc(timestamp)) %>%
    mutate(group = cumsum(!is.na(heart_rate))) %>%
    group_by(group) %>%
    mutate(aft_weight = n():1) %>%
    ungroup() %>%
    arrange(timestamp) %>%
    mutate(heart_rate = ifelse(is.na(heart_rate), (bef*bef_weight + aft*aft_weight) / (bef_weight + aft_weight), heart_rate))
    
  }

  
  hr_data <- hr_data %>%
    mutate(heart_rate = 60000 / heart_rate) %>%
    mutate(tp = 1:n()) %>% 
    dplyr::filter(tp > 600)
  

  
  tryCatch({

    y_true <- hr_data$heart_rate 
    x <- as.numeric(hr_data$timestamp)
    
    prop <- .5
    nit <- 10
    lambdas <- numeric(nit)
    for (i in 1:nit) {
        
      rem <- sort(sample(1:length(x), round(prop*length(x))))
      lambdas[i] <- optimise(objective_ns, interval = c(1, 1000), subset = rem, series = arma_garch_series, x = x)$minimum
        
    }
      
    lcs <- mean(lambdas)

    
    ## res <- gsm(hr_data$heart_rate ~ hr_data$tp, method = "GCV", control = list(maxit = 100, epsilon = 1e-9))
    res <- lm(arma_garch_series ~ ns(x, lcs))
    
    vals <- as.numeric(fitted(res))
    
    hr_data$vals <- vals
    hr_data$centered_hr <- hr_data$heart_rate - vals
    
    plot.ts(hr_data$centered_hr)
    
    ## model sigma t
    sfit <- ugarchfit(spec = ugarchspec(mean.model = list(armaOrder = c(0, 0)),
                                             variance.model = list(garchOrder = c(1, 1))), data = hr_data$centered_hr)
    
    ys <- sfit@fit$sigma
    xs <- 60000 / hr_data$heart_rate
    
    dat_pred <- bind_cols(ys, xs)
    colnames(dat_pred) <- c("Y", "X")
    
    mod <- lm(log(Y) ~ X, dat_pred)
    
    aerobic[i] <- exp(predict(mod, newdata = data.frame(X = 140)))
    dates[i] <- as.numeric(date(min(hr_data$timestamp)))
      
    },
           error = function(e) {
             print(i)
             print("Error")
             dates[i] <- NA
            }
  )
  
  i <- i + 1
  
}

saved <- list(dates, aerobic)

dates <- saved[[1]]
aerobic <- saved[[2]]

dates <- dates[!(aerobic == 0)]
act_type <- act_type[!(aerobic == 0)]
aerobic <- aerobic[!(aerobic == 0)]

plot(dates, aerobic, type = "l")
abline(lm(aerobic[act_type != "Ride"] ~ dates[act_type != "Ride"]), col = "red")

plot(dates, aerobic, type = "p", col = c("blue", "red")[1 + (act_type == "Ride")])
```



